The Cost of Exclusion: Political Discrimination and Trust Deficits in the European Union

### 2025-07-14 article/main.tex
new abstract

Simultaneously, ideological divergence can give rise to political dissonance—the psychological and emotional tension experienced by individuals who perceive a misalignment between their political identity and the dominant political climate or institutional stance \parencite{lelkes_limits_2017, federico_ideological_2012}.

Ideological divergence is exacerbated in polarized societies where outgroup hostility, moral condemnation, and dehumanization are common \parencite{iyengar_fear_2015}. This intensifies the emotional toll of political discrimination and heightens perceptions of institutional betrayal \parencite{hetherington_why_2015}. In turn, this reduces trust and discourages political participation \parencite{michelson_corrosive_2003, lelkes_limits_2017}.

>>>>

In the context of rising polarization, growing attention has been directed toward the increasing separation of ideological positions within the population—a phenomenon commonly referred to as ideological divergence. This concept captures the widening distance between the ideological stances of different groups, which can deepen societal divisions and reinforce partisan conflict. Ideological divergence often gives rise to psychological and emotional strain among individuals who perceive a misalignment between their own political identity and the dominant political climate or institutional orientation \parencite{lelkes_limits_2017, federico_ideological_2012}.

These effects are magnified in polarized societies, where out-group hostility, moral condemnation, and even dehumanization are prevalent \parencite{iyengar_fear_2015}. In such environments, political discrimination carries a heavier emotional burden, intensifying feelings of exclusion and perceptions of institutional betrayal \parencite{hetherington_why_2015}. As a result, ideological divergence not only undermines trust in institutions but also discourages political engagement and civic participation \parencite{michelson_corrosive_2003, lelkes_limits_2017}.
